---
title: "BMEG 310 MUTATION ANALYSIS ITERATION #1"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pathways_1 <- "/Users/ghazalfallahpour/Desktop/bmeg\ 310\ final\ proj\ data/data_clinical_patient.txt"
pathways_2 <- "/Users/ghazalfallahpour/Desktop/bmeg\ 310\ final\ proj\ data/data_mutations.txt"

clinical_data <- read.delim(pathways_1)
mutation_data <- read.delim(pathways_2)
```
 
## Preprocessing Data
```{r}

# Load the 'dplyr' library for data manipulation
library(dplyr)
library(ggplot2)
library("TCGAbiolinks")
library("survival")
library("survminer")
library("ggplot2")


# Clean the 'Tumor_Sample_Barcode' column in the mutation_data dataframe
mutation_data$Tumor_Sample_Barcode <- paste0(substr(mutation_data$Tumor_Sample_Barcode, 1, 12))

# Filter rows in mutation_data based on matching 'Tumor_Sample_Barcode' values in clinical_data
mutation_data <- mutation_data %>%
  filter(Tumor_Sample_Barcode %in% clinical_data$X.Patient.Identifier)

# Filter rows in clinical_data based on matching 'X.Patient.Identifier' values in mutation_data
clinical_data <- clinical_data %>%
  filter(X.Patient.Identifier %in% mutation_data$Tumor_Sample_Barcode)

# Rename the 'X.Patient.Identifier' column to 'Tumor_Sample_Barcode' in clinical_data
names(clinical_data)[names(clinical_data) == "X.Patient.Identifier"] <- "Tumor_Sample_Barcode"

# Extract the first two characters of 'American.Joint.Committee.on.Cancer.Tumor.Stage.Code'
clinical_data$American.Joint.Committee.on.Cancer.Tumor.Stage.Code <-
  substr(clinical_data$American.Joint.Committee.on.Cancer.Tumor.Stage.Code, 1, 2)

# Extract only the first character of 'Diagnosis.Age'
clinical_data$Diagnosis.Age <- substr(clinical_data$Diagnosis.Age, 1, 1)
```



## Data Visualization
```{r}
# Create a data frame 'hugo' with counts of unique values in 'Hugo_Symbol' column 
#using 'table'
hugo <- as.data.frame(table(mutation_data$Hugo_Symbol))

# Create a data frame 'var.class' with counts of unique values in 
#'Variant_Classification' column using 'table'
var.class <- as.data.frame(table(mutation_data$Variant_Classification))

# Create a bar plot using ggplot, displaying the frequency of each variant classification
ggplot(data=var.class, aes(x=Var1, y=Freq)) +
  geom_col() +  # Use geom_col() for a bar plot
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Adjust x-axis text
  #for better readability
  labs(x = "Variant Classification", y = "Frequency of Mutations")  # Label the axes

```

## First iteration of filtering 
```{r}

variants_to_remove <- c("Missense_Mutation", "Silent")
filtered_mutations <- mutation_data[!grepl(paste(variants_to_remove, collapse="|"),
                                           mutation_data$Variant_Classification), ]

#variants_to_keep <- c("stop_gained", "frameshift_variant","missense_variant",
#3"inframe_deletion")
#filtered_mutations <- mutation_data[grepl(paste(variants_to_keep, collapse="|"),
#mutation_data$Consequence), ]

#variants_to_keep <- c("HIGH", "MODERATE")
#filtered_mutations <- mutation_data[grepl(paste(variants_to_keep, collapse="|"),
#mutation_data$IMPACT), ]

#variants_to_keep <- c("HIGH")
#filtered_mutations <- mutation_data[grepl(paste(variants_to_keep, collapse="|"),
#mutation_data$IMPACT), ]

```


```{r}
# Create a data frame 'var.class2' with counts of unique values in 'VARIANT_CLASS'
#column using 'table'
var.class2 <- as.data.frame(table(filtered_mutations$VARIANT_CLASS))

# Create a bar plot using ggplot, displaying the frequency of each variant classification
# Use 'fill=Var1' to color bars based on the levels of the 'Var1' (variant classification) variable
ggplot(data=var.class2, aes(x=Var1, y=Freq)) +
  geom_col(aes(fill=Var1))  # Use geom_col() for a bar plot with fill color
```

```{r}
library(ggplot2)

# Create a data frame 'sample.name' with counts of unique values in 'Tumor_Sample_Barcode' 
#column using 'table'
sample.name <- as.data.frame(table(filtered_mutations$Tumor_Sample_Barcode))

# Create a data frame 'hugo' with counts of unique values in 'Hugo_Symbol' column using 'table'
hugo <- as.data.frame(table(filtered_mutations$Hugo_Symbol))

# Order the 'hugo' data frame based on the frequency of occurrences in descending order
hugo.ordered <- hugo[order(-hugo$Freq),]

# Create a bar plot using ggplot, displaying the top 25 most frequent genes
ggplot(data=hugo.ordered[1:25,], aes(x=Var1, y=Freq)) +
  geom_col() +  # Use geom_col() for a bar plot
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Adjust x-axis text for better 
#readability
  scale_x_discrete(limits = hugo.ordered[1:25,]$Var1) +  # Set x-axis limits for the 
#top 25 genes
  
  labs(x = "Hugo Code of Genes", y = "Frequency of Mutations in Genes")  # Label the axes
  

```

## Making the Onco Plot

```{r}

#Creating the Onco plot Matrix (Tutorial Code)

cnv_events = unique(filtered_mutations$Variant_Classification)
oncomat = reshape2::dcast(
  data = filtered_mutations,
  formula = Hugo_Symbol ~ Tumor_Sample_Barcode,
  fun.aggregate = function(x, cnv = cnv_events) {
    x = as.character(x) # >= 2 same/distinct variant classification = Multi_Hit
    xad = x[x %in% cnv]
    xvc = x[!x %in% cnv]
    
    if (length(xvc) > 0) {
      xvc = ifelse(test = length(xvc) > 1,
                   yes = 'Multi_Hit',
                   no = xvc)
    }
    
    x = ifelse(
      test = length(xad) > 0,
      yes = paste(xad, xvc, sep = ';'),
      no = xvc
    )
    x = gsub(pattern = ';$',
             replacement = '',
             x = x)
    x = gsub(pattern = '^;',
             replacement = '',
             x = x)
    return(x)
  },
  value.var = 'Variant_Classification',
  fill = '',
  drop = FALSE
)



# Set row names in the 'oncomat' dataframe to the values in the 'Hugo_Symbol' column
rownames(oncomat) = oncomat$Hugo_Symbol

# Remove the 'Hugo_Symbol' column from the 'oncomat' dataframe
oncomat <- oncomat[,-1]

# Order the 'oncomat' dataframe based on the frequency of occurrences in descending order
oncomat.ordered <- oncomat[order(-hugo$Freq),]

# Create a binary matrix 'mat' indicating presence (1) or absence (0) of mutations
mat <- oncomat.ordered
mat[mat!=""] <- 1  # Set non-empty entries to 1
mat[mat==""] <- 0  # Set empty entries to 0
mat <- apply(mat, 2 ,as.numeric)  # Convert the matrix to numeric values
mat <- as.matrix(mat)

# Set row names of 'mat' to the row names of 'oncomat.ordered'
rownames(mat)  <-  row.names(oncomat.ordered)
```

## Developing the Heatmap


```{r}
# Load the 'pheatmap' library for creating heatmap visualizations
library(pheatmap)

# Extract the first four rows of the binary matrix 'mat' for visualization 
#(iterations of 2,3,4,7 included)
#reduce.mat <- mat[1:2,]
#reduce.mat <- mat[1:3,]
#reduce.mat <- mat[1:4,]
reduce.mat <- mat[1:7,]



# Create a heatmap using pheatmap, without clustering rows, and without displaying column names
res <- pheatmap(reduce.mat,
                cluster_rows = F,  # Do not cluster rows
                show_colnames = FALSE)  # Do not display column names
```

## Clustering


```{r}
# Assign cluster labels to samples using cutree on the column dendrogram from the heatmap 
#(iterations of 2 and 3 included)
cluster <- as.data.frame(cutree(res$tree_col, k = 4))
#cluster <- as.data.frame(cutree(res$tree_col, k = 3))
#cluster <- as.data.frame(cutree(res$tree_col, k = 2))

# Initialize a new column 'cluster_group' in the 'clinical_data' dataframe with default value 1
clinical_data$cluster_group <- 1

# Loop through each row in the 'clinical_data' dataframe
for (i in 1:nrow(clinical_data)) {
  
    # Extract the Tumor_Sample_Barcode for the current row
    patient_id <- clinical_data$Tumor_Sample_Barcode[i]

    # Loop through each row in the 'cluster' dataframe
    for (m in 1:nrow(cluster)) {
      
      # Check if the Tumor_Sample_Barcode in 'cluster' matches the current 
      #Tumor_Sample_Barcode in 'clinical_data'
      if (rownames(cluster)[m] == patient_id) {
        # Assign the cluster label to the 'cluster_group' column in 'clinical_data'
        clinical_data$cluster_group[i] <- cluster[m, 1]
      }
      
    }
}

# Display the table of cluster labels and their counts
table(cluster)


```



```{r}
#choosing which clusters to continue survival analysis with
cluster_group_to_not_include_1 <- 1
cluster_group_to_not_include_2 <- 4


clinical_data1 <- subset(clinical_data, cluster_group != cluster_group_to_not_include_1 &
cluster_group != cluster_group_to_not_include_2)
```

## Survival Analysis 


```{r}
# Create a subset dataframe 'clin_df' from 'clinical_data1' with selected columns
clin_df = clinical_data1[
  c("Diagnosis.Age",
    "Disease.specific.Survival.status",
    "Overall.Survival..Months.",
    "Last.Communication.Contact.from.Initial.Pathologic.Diagnosis.Date",
    "Sex",
    "American.Joint.Committee.on.Cancer.Tumor.Stage.Code",
    "cluster_group"
  )]


# Create a binary column 'deceased' indicating whether the patient is dead with tumor
clin_df$deceased = clin_df$Disease.specific.Survival.status == "1:DEAD WITH TUMOR"

# Convert the date variable to numeric and scale to months
clin_df$Last.Communication.Contact.from.Initial.Pathologic.Diagnosis.Date <-
  as.numeric(clin_df$Last.Communication.Contact.from.Initial.Pathologic.Diagnosis.Date)
clin_df$Last.Communication.Contact.from.Initial.Pathologic.Diagnosis.Date <-
  cl
